---
title: "SDSS garden data demo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

These are the libraries you'll need:

```{r libraries}
library(tidyverse)         # for graphing and data cleaning
library(lubridate)         # for working with dates
# library(remotes)        # for installing package from GitHub
# remotes::install_github("llendway/gardenR") # run if package is not already installed
library(gardenR)           # for Lisa's garden data
```


# Data Visualization

The first place students meet the data is in data visualization. At first, I provide some of the data manipulation code. I will go through some of the examples I use in my class.

### Histogram

Histogram showing distribution of tomato harvests, in grams.

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("tomatoes")) %>% 
# I would provide the code for everything prior to the plot.
  ggplot(aes(x = weight)) +
  geom_histogram()
```

They learn to do some customizing of their plots.

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("tomatoes")) %>% 
# I would provide the code for everything prior to the plot.
  ggplot(aes(x = weight)) +
  geom_histogram(fill = "tomato") +
  geom_vline(aes(xintercept = mean(weight))) +
  labs(x = "Weight (grams)",
       y = NULL,
       title = "The average tomato harvest is just over 500 grams") +
  theme_minimal()
```


### Bar graph

Bar graph showing how many harvests of each vegetable. We'll improve this graph further later.

```{r}
garden_harvest %>% 
  ggplot(aes(y = vegetable)) +
  geom_bar() +
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "The number of harvests of each vegetable",
       x = NULL,
       y = NULL)
```


### Line graph

Line graph showing the daily tomato harvest in pounds on dates when tomatoes were harvested.

```{r}
# I provide this code:
tomatoes_wt_date <- garden_harvest %>% 
  filter(vegetable == "tomatoes") %>% 
  group_by(date) %>% 
  summarize(daily_wt_lb = sum(weight)*0.00220462)

# Students create this graph:
tomatoes_wt_date %>% 
  ggplot(aes(x = date, y = daily_wt_lb)) +
  geom_line() +
  labs(title = "Daily tomato harvest (lb)",
       x = NULL,
       y = NULL)
```


### Boxplot

Comparing the distributions of harvest weights in grams for a subset of vegetables. We'll improve this plot later, too.

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("lettuce", "tomatoes", "beans")) %>% 
  ggplot(aes(x = weight,
             y = vegetable)) +
  geom_boxplot() +
  labs(title = "Tomato harvests are much larger than lettuce harvests",
       x = "Weight (g)",
       y = NULL) +
  theme(plot.title.position = "plot")
```

## Data manipulation

Next, we move on to data manipulation. By this time, students have already seen some examples of this in code I wrote for them.

### Filtering

Filtering the vegetable types to lettuce, tomatoes, and beans.

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("lettuce", "tomatoes", "beans"))
```

And we can filter using multiple criteria. Filter the vegetables to tomatoes and cucumbers for dates after 08-10-2020

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("tomatoes", "cucumber"),
         date >= ymd("2020-08-10"))
```

### Adding new variables

Create a variable called `wt_lbs` that translates the weight in grams to weight in pounds. And, use the variable we just created to create another new variable called `over_1_lb` that is a logical variable describing whether the harvest is over 1 pound.

```{r}
garden_harvest %>% 
  mutate(wt_lbs = weight * 0.00220462,
         over_1_lb = wt_lbs > 1)
```

### Summarizing data for each category

For tomatoes, find the total weight in pounds harvested on each date they were harvested.

```{r}
tomatoes_wt_date <- garden_harvest %>% 
  filter(vegetable == "tomatoes") %>% 
  group_by(date) %>% 
  summarize(daily_wt_lb = sum(weight)*0.00220462)

tomatoes_wt_date
```

## More advanced data manipulation

Students continue to expand their data wrangling/preparing skills.

### String manipulation

They learn to work with strings. Going back to the bar graph showing how many harvests of each vegetable. We'll make the vegetables start with capital letters.

```{r}
garden_harvest %>% 
  mutate(vegetable = str_to_title(vegetable)) %>% 
  ggplot(aes(y = vegetable)) +
  geom_bar() +
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "The number of harvests of each vegetable",
       x = NULL,
       y = NULL)
```

### Reordering categorical data

They learn to reorder levels within the variable to make the plots more informative. 

```{r}
garden_harvest %>% 
  mutate(vegetable = str_to_title(vegetable)) %>% 
  ggplot(aes(y = fct_rev(fct_infreq(vegetable)))) +
  geom_bar() +
  scale_x_continuous(expand = c(0,0)) +
  labs(title = "The number of harvests of each vegetable",
       x = NULL,
       y = NULL)
```

### Joining datasets

I introduce another dataset from my garden, `garden_planting`. This dataset describes when and where I planted each vegetable variety. 

```{r}
data("garden_planting")
```

I would like to investigate the question: how many days from the date the variety of vegetable was planted did I get my first harvest?

I am going to summarize the `garden_planting` dataset first. For each vegetable and variety, I will find the minimum date. I will save that as a new dataset called `plant_min_date`.

```{r}
plant_min_date <- garden_planting %>% 
  group_by(vegetable, variety) %>% 
  summarize(first_plant_date = min(date))

plant_min_date
```

Now let's join this dataset to the `garden_harvest` data. (You might notice missing values at the end for varieties that didn't have a match in the garden_planting dataset).

```{r}
garden_harvest %>% 
  left_join(plant_min_date,
            by = c("vegetable", "variety")) %>% 
  mutate(date_diff = date - first_plant_date) %>% 
  group_by(vegetable, variety) %>% 
  summarize(days_to_first_harvest = min(date_diff)) %>% 
  arrange(days_to_first_harvest)
```

### Pivoting/transposing datasets

Here I have summarized the data for a tomatoes, peas, beans, and lettuce to find the total weight harvested on each day of the week. This table is pretty long to look through. 

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("tomatoes", "peas", "beans", "lettuce")) %>% 
  mutate(day = wday(date, label = TRUE)) %>% 
  group_by(vegetable, day) %>% 
  summarize(total_weight = sum(weight))
```

For looking at the results, I would rather have a column for each vegetable. Keep in mind, this is just for *viewing* the results. I'm assuming I won't be doing further analysis. If I want to keep playing with the data, it's better to leave it in its current form.

```{r}
garden_harvest %>% 
  filter(vegetable %in% c("tomatoes", "peas", "beans", "lettuce")) %>% 
  mutate(day = wday(date, label = TRUE)) %>% 
  group_by(vegetable, day) %>% 
  summarize(total_weight = sum(weight)) %>% 
  pivot_wider(names_from = vegetable,
              values_from = total_weight)

```

## And more!!

Students also use these datasets when learning how to make animated plots with `gganimate` and creating interactive shiny apps.






